Data Exploring, Cleaning, and Wrangling

### 2023-05-05 code/analysis_code/statistical_analysis.qmd
Part 5 of project
cleaned up manuscript, removed extra text, fixed graphs/plots
Final project commit

---
title: "Statistical Analysis"
author: "Sara Benist"
date: "2023-03-17"
output: html_document
---

@@

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The processed data is used for model fitting.

@@

```{r, loaddata}
```
# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus on linear models of the outcomes TB incidence, TB mortality, and TB prevalence. I will be fitting two models using the full data set and only the high burden countries.

## Linear model fitting for TB Incidence

```{r, fitmodel1}
```
The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Since many of the countries in the full data set have very low incidence, the coefficients for the intercept and subgroup variable are much lower. The model has more data points with low values.

### Prediction model for TB Incidence
We can also try to predict the approximate TB incidence for a group. For this example, we will include subgroup and setting as predictors.  
```{r, pred1}
```
With the prediction model including setting, there are a lot of variables in the model. The prediction would have a 1 for the country of interest and 0 for all other setting variables. For example, the prediction for a male in Central African Republic would appear as `TB incidence = 126(1)+189(1)`.

As an example, I will predict the incidence rate for males and females in India. The code for prediction modeling is based off of the `tidymodels` guidance. The plot shows the incidence rate and confidence intervals for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) in India. This can be repeated with the high burden countries.

## Linear model fitting for TB Mortality

```{r, fitmodel2}
```
For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`.

## Linear model fitting for TB Prevalence

```{r, fitmodel3}
```
For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`. The difference between the high burden and full data models is lower than the other outcomes, most likely due to the long duration of TB or less data points available with the full data model.

## Linear model fitting for BCG Coverage
Now we can look at some of the other indicators in the data set. I will only be looking at the indicators that had variation between subgroups. 
```{r, fitmodel4}
```
The BCG coverage has 12 variables that produce a linear equation of `BCG coverage % = 88.4 + 0.363(Male) -10.6(no education) + 0.169(primary education) + 6.66(higher education) - 6.46(quintile 1) - 1.6(quintile 2) +1.06(quintile 3) + 4.50(quintile 4) + 7.01(quintile 5) - 2.12(rural) + 5.10(Urban)`. The full model has an equation of `BCG coverage % = 88.3 + 0.245(Male) -9.44(no education) + 0.958(primary education) + 6.42(higher education) - 6.40(quintile 1) - 1.68(quintile 2) + 0.773(quintile 3) + 3.84(quintile 4) + 6.41(quintile 5) - 2.08(rural) + 4.75(Urban)`. The two models are overall pretty similar and can be used in the same way as the prediction model (entering in 1 or 0 depending on categorization).

## Linear model fitting for Catastrophic costs

```{r, fitmodel5}
```
The high burden model has an equation of `% catastrophic costs = 87.3 - 40.4(drug-suseptible TB) - 18.6(quintile 1) - 38.8(quintile 2) - 44.6(quintile 3) - 47.4(quintile 4) - 58.5(quintile 5)`. The full model data has an equation of `% catastrophic costs = 83.5 - 35.1(drug-suseptible TB) - 11.1(quintile 1) - 30.9(quintile 2) - 38.4(quintile 3) - 41.4(quintile 4) - 54.9(quintile 5)`.
## Linear model fitting for case detection rate

```{r, fitmodel6}
```
For high burden countries, the equation is `case detection rate % = 37.9 + 19.1(15+ or older) + 16.7(female) + 16.4(male)`. The full data model has an equation of `case detection rate % = 43.8 + 18.9(15+ or older) + 15.3(female) + 16.9(male)`.

## Linear model fitting for prevalence to notification ratio

```{r, fitmodel7}
```
The high burden model has an equation of `prevalence to notification ratio in years = 1.73 + 0.664(male)`, and the full model has an equation of `prevalence to notification ratio in years = 1.73 + 0.701(male)`.

## Linear model fitting for attitudes

```{r, fitmodel8}
```
The attitude question regarding TB was about the desire to keep a diagnosis hidden. Since males are more likely to be diagnosed with TB, I focused on the male attitude and knowledge. For high burden countries, the equation is `% wanting to keep TB secret = 27.6 - 5.27(20-34 years) - 8.21(35-49 years) - 3.53(no education) - 2.71(primary education) - 7.23(higher education) - 5.01(quintile 1) - 5.23(quintile 2) - 4.97(quintile 3) - 4.38(quintile 4) - 5.86(quintile 5) - 4.91(rural) - 6.27(urban)`. For the full model, the equation is `% wanting to keep TB secret = 24.7 - 4.94(20-34 years) - 8.14(35-49 years) - 4.32(no education) - 3.08(primary education) - 6.41(higher education) - 4.44(quintile 1) - 5.24(quintile 2) - 5.18(quintile 3) - 4.59(quintile 4) - 6.20(quintile 5) - 4.66(rural) - 5.99(urban)`.

## Linear model fitting for knowledge

```{r, fitmodel9}
```
The knowledge question asked about if TB is spread through cough. We will again focus on male knowledge.

The high burden countries had an equation of `% responding TB is spread through cough = 65.9 + 4.83(20-34 years) + 6.19(35-49 years) - 9.91(no education) - 2.81(primary education) + 12.6(higher education) - 6.01(quintile 1) - 0.464(quintile 2) + 2.32(quintile 3) + 7.50(quintile 4) + 14.6(quintile 5) + 0.419(rural) + 11.8(urban)`. The full model had an equation of `% responding TB is spread through cough = 64.6 + 6.17(20-34 years) + 7.29(35-49 years) - 8.55(no education) - 1.66(primary education) + 12.8(higher education) - 3.88(quintile 1) + 0.028(quintile 2) + 3.03(quintile 3) + 8.18(quintile 4) + 15.8(quintile 5) + 1.28(rural) + 12.8(urban)`.

I would like to try and create a dataset using the high burden countries in order to predict the main outcomes of TB incidence and mortality from the indicators (BCG Coverage, knowledge, attitudes, etc). Right now, the data is not set up to complete this analysis, but I will try to find a way to do this.

>>>>

---
title: "Decision Tree Statistical Analysis"
author: "Sara Benist"
date: "2023-04-06"
output: html_document
---

@@

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The best way to view the output of the script is to render the qmd and view the html document. The processed data is used for model fitting.

*Note: this file takes approximately 10 minutes to render*

@@

```{r, loaddata}
```

# Basic modelling
For some basic modeling, I will be fitting linear models to the TB outcomes with the subgroups as predictors. Subgroups are the main difference between populations other than country which is why I chose these as the bivariate models. This will offer another view of the data. The prediction model also offers a way to explore the data by submitting a sample prediction of TB incidence for females and males in Central African Republic.
```{r, basicmodels}
```
The models indicate that subgroup has a significant influence on the outcomes since the estimates for the subgroup (being Male primarily or living in rural areas) increases the incidence, mortality, or prevalence. This is expected from the exploratory analysis.

The biggest issue with the linear modeling is the assumption that the relationship between predictors and outcomes is linear, and the model cannot be tuned and replicated.

# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will split the data into testing and training sets and set up the cross-validation folds. The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.

```{r, splitdata}
```

I will be tuning decision tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the other burden set. In the `boosted_statistical_analysis.qmd` file, a boosted tree model for each outcome is created for comparison. In order to be clear when saving the resulting tables and figures, the other burden outcomes from the decision tree model are labeled as `obdt` and the high burden as `hbdt`. The testing data will be fit to the best performing models.

## Decision tree model for TB Incidence
```{r, obdtincidencemodel}
```
The best tree model for TB incidence using the other burden countries has a depth of 4 with an RMSE of 51.6. This is a very low level of performance in predicting TB incidence (ideally would be closer to 1). The fitted model to the training data has a RMSE of 29.8.

```{r, hbdtincidencemodel}
```
The best tree model for TB incidence using the high burden countries has a depth of 4 with an RMSE of 159. This is also a low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 81.3 which was much higher compared to the full dataset.

## Decision tree model for TB Mortality
```{r, obdtmortalitymodel}
```
The best tree model for TB mortality using the other burden has a depth of 8 with an RMSE of 8.85. The fitted model to the training data has a RMSE of 7.21.

```{r, hbdtmortalitymodel}
```
The best tree model for TB mortality using the high burden countries has a depth of 4 with an RMSE of 20.4. The fitted model to the training data has a RMSE of 10.3.

## Decision tree model for TB Prevalence

```{r, obdtprevalencemodel}
```
The best tree model for TB prevalence using the other burden countries has a depth of 1 with an RMSE of 294. This is a very low level of performance in predicting TB prevalence. The fitted model to the training data has a RMSE of 240.

```{r, hbdtprevalencemodel}
```
The best tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 322. This is a very low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 254.

The models were overall not a good predictor of TB outcomes. This is possibly due to the large variance in the outcomes as well as the summarizing required to de-aggregate the data and predict outcomes. Country was overwhelmingly the most important predictor.

## Summarise models
Overall, the models were not great predictors of TB outcomes. The mortality performed the best, and all high burden models had lower RMSE values by about half of the other burden models. 
In the `boosted_statistical_analyisis.qmd` file, I created boosted tree models as a comparison. The boosted tree models should perform better than a decision tree model. 
```{r, summary}
```

### 2023-05-05 code/processing_code/processingfile.qmd
Final project commit

I will also be looking specifically at high TB burden countries. The high burden countries list is updated by the WHO and can be found [here](https://www.who.int/news/item/17-06-2021-who-releases-new-global-lists-of-high-burden-countries-for-tb-hiv-associated-tb-and-drug-resistant-tb). I will subset these countries in the `highburden` object.

@@

The `wide_data` had 7473 rows and 20 columns. The `highburden` object has 3761 rows and 20 columns.

@@

>>>>

I will also be looking specifically at high TB burden countries. The high burden countries list is updated by the WHO and can be found [here](https://www.who.int/news/item/17-06-2021-who-releases-new-global-lists-of-high-burden-countries-for-tb-hiv-associated-tb-and-drug-resistant-tb). I will subset these countries in the `highburden` object. The countries with low or mid level burden will be sorted into `otherburden`.

@@

The `wide_data` had 7473 rows and 20 columns with 194 countries. The `highburden` object has 3761 rows and 20 columns with 49 countries. The `otherburden` object has 3712 rows and 20 columns with 145 countries.

@@

# Modelling setup

The final analysis will be predicting tuberculosis outcomes from the indicators. Due to the nature of the data, each indicator must be summarized for males and females (aggregate data can measure males or someone in the lowest economic quintile but cannot predict the indicator measure for males in the lowest economic quintile). To set up the data for modelling, the data is summarized by sex and indicator for each country in the high burden and other burden data sets. The incidence and mortality outcomes consist of indicators with sex as a dimension, and the prevalence outcome consists of rural or urban subgroups.

```{r, modelsetup}
```
